The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
proof: caterpillar distance = distance if max

A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

\begin{lemma}
    Let $T$ and $R$ be two trees on $n+1$ taxa.
    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
    In particular, this holds for a shortest path $p$ from $T$ to $R$.
    Furthermore, deleting taxon $n+1$ and its parent $p_{n+1}$ from all trees on $p$ results in a path $p|_n$ from $T|_n$ to $R|_n$.
    All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
    \todo{Is this proof sufficient? Otherwise: induction}
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    % \end{enumerate}
\end{proof}

@@

\begin{lemma}
    Let $T$ and $R$ be two caterpillar trees.
    $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$. \todo{define $d_c$}
    \label{lemma:caterpillar_dist=diameter}
\end{lemma}

    % \todo{At the top of $R$ it could also be $2,1$, or $t=1$}
    %
    % \begin{figure}[H]
    %     \centering
    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
    %     \caption{Caterpillar trees $T$ and $R$ with distance $d(T,R) = \Delta(\rnniu)$}
    %     \label{fig:caterpillar_max_distance}
    % \end{figure}

\begin{proof}
    Let $d_c(T,R)$ denote the length of a shortest caterpillar path from $T$ to $R$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $R$.
    We need to prove that $d(T,R) = \Delta(\rnniu)$ if, and only if, $d_c(T,R) = \Delta(\rnniu)$.

    Let $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu)$.
    We can compute a  shortest caterpillar path from $T$ to $R$ using Bubble Sort.
    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.

    Let now $T$ and $R$ have caterpillar distance $d_c(T,R) = \Delta(\rnniu)$.
    We prove $d(T,R) = d_c(T,R)$ by induction on the number of taxa $n$ of $T$ and $R$:
    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
    Otherwise, we re-label $T$ and $R$.
    Let $R$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.

    Induction basis: $n=3$\\
    All trees have distance $1$ and are caterpillar trees.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
    When applying the Bubble Sort algorithm to compute a path between $R$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $R$.
    So we can conclude that taxon $n+1$ must be in the cherry of $R$.
    Let us consider the trees $T|_n$ to $R|_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    When running Bubble Sort to compute a path from $T|_n$ to $R|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $R$, except for the $n-1$ ones that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T|_n, R|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $R|_n$ and know that $d(T|_n,R|_n) = \frac{(n-1)(n-2)}{2}$.

    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    We also know $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = n-1$
    If we assume that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    So it would be $d(T|_n, R|_n) < \Delta(\rnniu_n)$.
    Since this is a contradiction to $d(T|_n, R|_n) = \Delta(\rnniu_n)$, it must be $d(T,R) = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.

@@

\begin{proof}
	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
    Therefore, the length of a shortest caterpillar path is \sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}.
\end{proof}

>>>>

A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

\begin{lemma}
    Let $T$ and $R$ be two trees on $n+1$ taxa.
    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
    In particular, this holds for a shortest path $p$ from $T$ to $R$.
    Furthermore, deleting taxon $n+1$ and its parent $p_{n+1}$ from all trees on $p$ results in a path $p|_n$ from $T|_n$ to $R|_n$.
    All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    \todo{Is this proof sufficient? Otherwise: induction}
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    % \end{enumerate}
\end{proof}

@@

\begin{lemma}
    Let $T$ and $R$ be two caterpillar trees.
    $T$ and $R$ have distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
    \label{lemma:caterpillar_dist=diameter}
\end{lemma}

\begin{proof}
    Let $T$ and $R$ be caterpillar trees.
    Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
    This easily follows from the modified Bubble Sort algorithm. \todo{name?}
    In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
    It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.

    For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
    We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
    Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.

    Induction basis: $n=3$\\
    All three trees have distance $1$ from another and are caterpillar trees, hence it is $d(T,R) = d_c(T,R)$ for all trees $T,R$ on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{n(n-1)}{2}$.\\
    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{(n-1)(n-2)}{2}$.
    Therefore, each of the $\frac{(n-1)(n-2)}{2}$ pairs of taxa that is considered when running modified Bubble Sort algorithm \todo{name} to compute a path $p$ from $R$ to $T$ exchanges.
    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
    Let us consider the trees $T|_n$ to $R|_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    Bubble Sort \todo{name} computes a path from $T|_n$ to $R|_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T|_n, R|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $R|_n$ and know that $d(T|_n,R|_n) = \frac{(n-1)(n-2)}{2}$.

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    We also know $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = n-1$
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T|_n, R|_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.

@@

\begin{proof}
	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
    Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
\end{proof}
